---
title: "Biostats 213 Midterm Project"
author: Khoa Tran 505029406
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Assume that your restaurant opens at 8am and closes at 5pm.  Customers arrive at a rate of 4/hr.  Between the hours of noon and 1 that rate increases to 6/hr.  Suppose you have 3 tellers. We have studied 3 types of queuing systems 1) 1 line with parallel servers.  2)  tandem service (In-N-out example) and 3) where each teller has their own line (grocery store).  Given three tellers which queuing system has the shortest expected waiting time for the customers?

## SIMULATE 3 QUEUING SYSTEMS:  

* Service completion time of server 1 is generated from `rgamma(1, 1, 10)`.  
* Service completion time of server 2 is generated from `rgamma(1, 1, 11)`. 
* Service completion time of server 3 is generated from `rgamma(1, 1, 12)`.
* Each system was simulated over 100 days.

### RESULTS:  
The wait time is shortest for system with 1 line at about 5.67 minutes. The wait time is longest for system with 3 tandem tellers at about 6.36 minutes. The wait time for system with 3 independent wait lines is in between, at about 5.99 minutes.  

* The average wait time for 1 line 3 tellers is 0.0945 (~5.67 mins) hour.  
* The average wait time for 3 tandem tellers is 0.1060 (~5.99 mins) hour.
* The average wait time for 3 tellers each with their independent line is 0.0999 (~6.36 mins) hour.  

### CODE

#### Functions used in the 3 simulations  

#### Function to change lambda based on time of the day

```{r}
lambda.t <- function(time){		# the banks open from 8am to 5pm, so we set 	
  if(time >= 4 && time <= 5){	# the time from 0 to 9 and the rate is 4/hr 
    lambda <- 6		# but increase to 6 between 4 and 5.
  } else { lambda <- 4 }
  list(lamt=lambda)
}
```

#### Function to generate arrival time based on previous arrival time and time t

```{r}
generate.Tt <- function(Ta, t) {
  Tt <- t- (1/lambda.t(t)$lamt)*log(runif(1))	# next arrival time
  
  if (Ta < 4) {		# Ta is less than 4
    Ta <- Tt
    if (Ta > 4) { Ta <- 4+((Ta-4)*4/6) }	#if greater than 4, change rate	
  } else if ( 4 <= Ta && Ta <= 5 ) {	# 4 <= Ta <= 5
    Ta <- Tt
    if (Ta > 5) { Ta <- 5+((Ta-5)*6/4) }	#if greater than 5 , change rate
  } else { 
    Ta <- Tt
  }
  
  if ( Ta > 9)  Ta <- 9999	# To insure that nobody can enter the system after 
  return(Ta)
}		
```

### 1) One line with three parallel servers

```{r}
set.seed(213)     #seed = 213
N <- 100
cust.wait <- numeric(N)
day <- 1
for(day in 1:N){
  t <- 0          #time starts at 0
  T <- 9          #store closed 9 hrs later
  A <- numeric()  #vector of arrival time of customer
  D <- numeric()  #vector departure time of customer
  Na <- 0         #number of arrival by time t
  Nd <- 0         #number of departure by time t 
  n <- 0          #number of customer in the store
  i1 <- 0         #index of the customer at server 1
  i2 <- 0         #index of the customer at server 2
  i3 <- 0         #index of the customer at server 3
  t1 <- 9999      #service completion time of server 1
  t2 <- 9999      #service completion time of server 2
  t3 <- 9999      #service completion time of server 3
  c1 <- 0         #number of customer served by server 1
  c2 <- 0         #number of customer served by server 2
  c3 <- 0         #number of customer served by server 3
  
  #Generate the first arrival time
  Ta <- t - (1/lambda.t(t)$lamt) * log(runif(1))
  
  #While the store is open
  while (t < T) {
    #Case 1: someone arrives before someone departs or the servers are all idle
    if (Ta == min(Ta, t1, t2, t3)) {
      t <- Ta
      Na <- Na + 1
      A[Na] <- t 
      Ta <- generate.Tt(Ta, t) #generate the next arrival time
      
      #if the store is empty
      #server 1 has priority, then server 2, then server 3
      if (n == 0 & t1 == 9999 & t2 == 9999 & t3 == 9999) {
        n <- 1
        i1 <- Na #server 1 has priority, then server 2, then server 3
        Y1 <- rgamma(1, 1, 10) #service period
        t1 <- t + Y1
      } else if (n == 1 & t1 != 9999 & t2 == 9999 & t3 == 9999) {
        #server 1 is busy, server 2 and 3 are idle
        n <- 2
        i2 <- Na  #server 2 has priority over server 3
        Y2 <- rgamma(1, 1, 11)
        t2 <- t + Y2
      } else if (n == 1 & t1 == 9999 & t2 != 9999 & t3 == 9999) {
        #server 2 is busy, server 1 and 3 are idle
        n <- 2
        i1 <- Na #go to server 1
        Y1 <- rgamma(1, 1, 10)
        t1 <- t + Y1
      } else if (n == 1 & t1 == 9999 & t2 == 9999 & t3 != 9999) {
        #server 3 is busy, server 1 and 2 are idle
        n <- 2
        i1 <- Na #go to server 1
        Y1 <- rgamma(1, 1, 10)
        t1 <- t + Y1
      }else if(n == 2 & t1 != 9999 & t2 != 9999 & t3 == 9999){
        #server 1 and 2 are busy, server 3 is idle
        n <- 3
        i3 <- Na #go to server 3
        Y3 <- rgamma(1, 1, 12)
        t3 <- t + Y3
      } else if (n == 2 & t1 != 9999 & t2 == 9999 & t3 != 9999) {
        #server 1 and 3 are busy, server 2 is idle
        n <- 3
        i2 <- Na #go to server 2
        Y2 <- rgamma(1, 1, 11)
        t2 <- t + Y2
      } else if (n == 2 & t1 == 9999 & t2 != 9999 & t3 != 9999) {
        #server 2 and 3 are busy, server 1 is idle
        n <- 3
        i1 <- Na #go to server 1
        Y1 <- rgamma(1, 1, 10)
        t1 <- t + Y1
        
        #else increase the line
      } else if (n > 2) {
        n <- n + 1
      }
    
    #Case 2: someone leaves server 1 before someone arrives
    } else if (t1 == min(Ta, t1, t2, t3)) {
      t <- t1
      c1 <- c1 + 1
      Nd <- Nd + 1
      D[i1] <- t
      #if there less than or = 3 customers, one leaves server 1 -> server 1 becomes idle
      if (n <= 3) {
        n <- n - 1
        i1 <- 0
        t1 <- 9999
      } else if (n > 3) {  #if more than 3 customers, server 1 serves next in line
        n <- n - 1
        m <- max(i1, i2, i3)
        i1 <- m + 1
        Y1 <- rgamma(1, 1, 10)
        t1 <- t + Y1
      }
      
    #Case 3: someone leaves server 2 before someone arrives 
    } else if (t2 == min(Ta, t1, t2, t3)) {
      t <- t2
      c2 <- c2 + 1
      Nd <- Nd + 1
      D[i2] <- t
      if (n <= 3) {
        n <- n - 1
        i2 <- 0
        t2 <- 9999
      } else if (n > 3) {
        n <- n - 1
        m <- max(i1, i2, i3)
        i2 <- m + 1
        Y2 <- rgamma(1, 1, 11)
        t2 <- t + Y2
      }
      
    #Case 4: someone leaves server 3 before someone arrives
    } else if (t3 == min(Ta, t1, t2, t3)) {
      t <- t3
      c3 <- c3 + 1
      Nd <- Nd + 1
      D[i3] <- t
      if (n <= 3) {
        n <- n - 1
        i3 <- 0
        t3 <- 9999
      } else if (n > 3) {
        n <- n - 1
        m <- max(i1, i2, i3)
        i3 <- m + 1
        Y3 <- rgamma(1, 1, 12)
        t3 <- t + Y3
      }
    }
    if (min(Ta, t1, t2, t3) >= T) break
  }
  
  #When time > 9, store closes but there are still customers
  if (min(Ta, t1, t2, t3) > T & n > 0) {
    while (n > 0) {
      #someone leaves server 1 before server 2 and 3
      if (t1 == min(t1, t2, t3)) {
        t <- t1
        Nd <- Nd + 1
        c1 <- c1 + 1
        D[i1] <- t
        if (n <= 3) {
          n <- n - 1
          i1 <- 0
          t1 <- 9999
        } else if(n > 3) {
          n <- n - 1
          m <- max(i1, i2, i3)
          i1 <- m + 1
          Y1 <- rgamma(1, 1, 10)
          t1 <- t + Y1
        }
        
      #someone leaves server 2 before server 1 and 3 
      } else if (t2 == min(t1, t2, t3)) {
        t <- t2
        Nd <- Nd + 1
        c2 <- c2 + 1
        D[i2] <- t
        if (n <= 3) {
          n <- n - 1
          i2 <- 0
          t2 <- 9999
        } else if (n > 3) {
          n <- n - 1
          m <- max(i1, i2, i3)
          i2 <- m + 1
          Y2 <- rgamma(1, 1, 11)
          t2 <- t + Y2
        }
        
      #someone leaves server 3 before server 1 and 2
      } else if (t3 == min(t1, t2, t3)) {
        t <- t3
        Nd <- Nd + 1
        c3 <- c3 + 1
        D[i3] <- t
        if (n <= 3) {
          n <- n - 1
          i3 <- 0
          t3 <- 9999
        } else if (n > 3) {
          n <- n - 1
          m <- max(i1, i2, i3)
          i3 <- m + 1
          Y3 <- rgamma(1, 1, 12)
          t3 <- t + Y3
        }
      }
    }
  }
  #collect average wait time
  cust.wait[day] <- mean(D - A)
  day <- day + 1
}

mean(cust.wait)
```

### 2) Tandem service (one line with three check out points)

```{r}
set.seed(213)
N <- 100
cust.wait2 <- numeric(N)
for (day in 1:N) {
  t <- 0                
  T <- 9                
  Na <- 0 
  Nd <- 0 
  n1 <- 0               #number of customer in line for teller 1 at time t
  n2 <- 0               #number of customer in line for teller 2 at time t
  n3 <- 0               #number of customer in line for teller 3 at time t
  t1 <- 9999            #service completion time at teller 1
  t2 <- 9999            #service completion time at teller 2
  t3 <- 9999            #service completion time at teller 3
  ANa.1 <- numeric()    #arrival time at server 2
  ANa.2 <- numeric()    #arrival time at server 2
  ANa.3 <- numeric()    #arrival time at server 3
  DNd <- numeric()      #departure time
  
  #generate the first arrival time
  Ta <- t - (1/lambda.t(t)$lamt)*log(runif(1))  
  
  repeat{
    if(Ta==9999 && Ta==t1 && t1==t2 && t2==t3) break
    ### CASE 1: someone arrives at server 1
    if (Ta == min(Ta,t1,t2,t3)) {
      t <- Ta
      Na <- Na + 1
      n1 <- n1 + 1
      Ta <- generate.Tt(Ta, t) # NOtice: the change of rate
      if (n1 == 1) {
        t1 <- t + rgamma(1, 1, 10) / 3    
      }
      ANa.1[Na] <- t
      
      
    ### CASE 2: someone leaves server 1
    } else if (t1 == min(Ta,t1,t2,t3)) { 
      t <- t1 
      n1 <- n1 - 1  
      n2 <- n2 + 1 
      if (n1 == 0) {
        t1 <- 9999
      } else {
        t1 <- t + rgamma(1, 1, 10) / 3
      }
      if (n2 == 1) {t2 <- t + rgamma(1, 1, 11) / 3}
      ANa.2[(Na-n1)] <- t
      
    ### CASE 3: someone leaves server 2
    } else if ( t2 == min(Ta,t1,t2,t3)) {  
      t <- t2
      n2 <- n2 - 1
      n3 <- n3 + 1
      if (n2 == 0) {
        t2 <- 9999
      } else {
        t2 <- t +rgamma(1, 1, 11) / 3 
      }
      if (n3 == 1){ t3 <- t + rgamma(1, 1, 12) / 3 }
      ANa.3[(Na-n1-n2)] <- t
      
      ### CASE 4: someone leaves server 3
    } else if ( t3 == min(Ta,t1,t2,t3)) {  
      t <- t3
      n3 <- n3 - 1
      Nd <- Nd + 1
      if (n3 == 0) {
        t3 <- 9999
      } else {
        t3 <- t + rgamma(1, 1, 12) / 3
      }
      DNd[(Nd)] <- t
    }
  }
  cust.wait2[day] <- mean(DNd-ANa.1)
  day <- day + 1
}

mean(cust.wait2)
```

### 3) Each teller has their own line (grocery store)  
* For this project, each line is treated as independent from each other.
* New lambda function, the rate of arrival is divided by 3 as one teller does not service all the customers

```{r}
lambda.t <- function (time) {		# the banks open from 8am to 5pm, so we set 	
  if (time >= 4 && time <= 5) {	# the time from 0 to 9 and the rate is 4/hr 
    lambda <- 6	/3	# but increase to 6 between 4 and 5.
  } else {lambda <- 4 / 3}
  list(lamt = lambda)
}
```

```{r}
set.seed(213)
N<-100

## QUEUE FOR TELLER 1

cust.wait1 <- numeric(N)
Tp1 <- numeric(N)
depart1 <-numeric(N)
Timepast1 <-numeric(N)
i<-0

for (i in 1:100) {		#starting the loop and initializing
  t<-0				#time starts at 0
  T<-9				#time is 8 hours
  A <- 0      #arrival time
  D <- 0      #departure time
  Na <-0      #number of arrivals
  Nd <-0      #number of departures
  n  <-0      #number of customer at time t
  Td <- 9999
  
  #generate arrival times
  Ta <- t - (1/lambda.t(t)$lamt)*log(runif(1))

  while(t<T)
  {
    #Case 1: somebody arrives before somebody departs and store is open
    if (Ta <= Td & Ta <= T)
    {
      t <- Ta				#updating time variable
      Na <- Na + 1			#add additional arrival at time ta
      n <- n + 1			#add additional customer
      A[Na] <- t			#output data (customer Na arrived at time t)
      Ta <- generate.Tt(Ta, t)

      if (n==1)		#if n=1 we need to generate Y and reset
      {		    #td=t+Y because system was empty and td=infty
        Y <- rgamma(1,1,10)
        Td <- t + Y
      }
    }
    
    #CASE 2: somebody departs before somebody arrives
    if (Td < Ta & Td <=T)
    {
      t <- Td
      n <- n-1
      Nd <- Nd + 1
      D[Nd] <-t
      if (n == 0) 
      {Td <- 9999 
      }
      if (n > 0)  #there is customer in queue
      {	
        Y <-  rgamma(1,1,10)
        Td <- t + Y
      }
    }
    
    #Case 3: store is closed but there is still customer
    if (min(Ta,Td) > T & n > 0)
    {
      while (n >0)
      {
        t <- Td
        n<- n-1
        Nd<- Nd + 1
        D[Nd] <-  t
        if (n >0)
        {
          Y <- rgamma(1,1,10)
          Td <-t + Y
        }
      }
    }
    
    #Case 4: close and no more customer
    if (min(Ta,Td) > T & n==0)
    {
      Tp<-max(t-T,0)
      Timepast1[i] <-Tp
    }
    if(min(Ta, Td) >= T) break
  }
  
  cust.wait1[i] <- sum(D-A)/Na
  depart1[i] <-t
  i <- i+1
}
mean(cust.wait1) #average wait time for teller 1

## QUEUE FOR TELLER 2

cust.wait2 <- numeric(N)
Tp2 <- numeric(N)
depart2 <-numeric(N)
Timepast2 <-numeric(N)
i<-0

for (i in 1:100) {		#starting the loop and initializing
  t<-0				#time starts at 0
  T<-9				#time is 8 hours
  A <- 0      #arrival time
  D <- 0      #departure time
  Na <-0      #number of arrivals
  Nd <-0      #number of departures
  n  <-0      #number of customer at time t
  Td <- 9999
  
  #generate arrival times
  Ta <- t - (1/lambda.t(t)$lamt)*log(runif(1))
  
  while(t<T)
  {
    #Case 1: somebody arrives before somebody departs and store is open
    if (Ta <= Td & Ta <= T)
    {
      t <- Ta				#updating time variable
      Na <- Na + 1			#add additional arrival at time ta
      n <- n + 1			#add additional customer
      A[Na] <- t			#output data (customer Na arrived at time t)
      Ta <- generate.Tt(Ta, t)
      
      if (n==1)		#if n=1 we need to generate Y and reset
      {		    #td=t+Y because system was empty and td=infty
        Y <- rgamma(1,1,11)
        Td <- t + Y
      }
    }
    
    #CASE 2: somebody departs before somebody arrives
    if (Td < Ta & Td <=T)
    {
      t <- Td
      n <- n-1
      Nd <- Nd + 1
      D[Nd] <-t
      if (n == 0) 
      {Td <- 9999 
      }
      if (n > 0)  #there is customer in queue
      {	
        Y <-  rgamma(1,1,11)
        Td <- t + Y
      }
    }
    
    #Case 3: store is closed but there is still customer
    if (min(Ta,Td) > T & n > 0)
    {
      while (n >0)
      {
        t <- Td
        n<- n-1
        Nd<- Nd + 1
        D[Nd] <-  t
        if (n >0)
        {
          Y <- rgamma(1,1,11)
          Td <-t + Y
        }
      }
    }
    
    #Case 4: close and no more customer
    if (min(Ta,Td) > T & n==0)
    {
      Tp<-max(t-T,0)
      Timepast2[i] <-Tp
    }
    if(min(Ta, Td) >= T) break
  }
  
  cust.wait2[i] <- sum(D-A)/Na
  depart2[i] <-t
  i <- i+1
}
mean(cust.wait2) #average wait time for teller 2

## QUEUE FOR TELLER 3

cust.wait3 <- numeric(N)
Tp3 <- numeric(N)
depart3 <-numeric(N)
Timepast3 <-numeric(N)
i<-0

for (i in 1:100) {		#starting the loop and initializing
  t<-0				#time starts at 0
  T<-9				#time is 8 hours
  A <- 0      #arrival time
  D <- 0      #departure time
  Na <-0      #number of arrivals
  Nd <-0      #number of departures
  n  <-0      #number of customer at time t
  Td <- 9999
  
  #generate arrival times
  Ta <- t - (1/lambda.t(t)$lamt)*log(runif(1))
  
  while(t<T)
  {
    #Case 1: somebody arrives before somebody departs and store is open
    if (Ta <= Td & Ta <= T)
    {
      t <- Ta				#updating time variable
      Na <- Na + 1			#add additional arrival at time ta
      n <- n + 1			#add additional customer
      A[Na] <- t			#output data (customer Na arrived at time t)
      Ta <- generate.Tt(Ta, t)
      
      if (n==1)		#if n=1 we need to generate Y and reset
      {		    #td=t+Y because system was empty and td=infty
        Y <- rgamma(1,1,12)
        Td <- t + Y
      }
    }
    
    #CASE 2: somebody departs before somebody arrives
    if (Td < Ta & Td <=T)
    {
      t <- Td
      n <- n-1
      Nd <- Nd + 1
      D[Nd] <-t
      if (n == 0) 
      {Td <- 9999 
      }
      if (n > 0)  #there is customer in queue
      {	
        Y <-  rgamma(1,1,12)
        Td <- t + Y
      }
    }
    
    #Case 3: store is closed but there is still customer
    if (min(Ta,Td) > T & n > 0)
    {
      while (n >0)
      {
        t <- Td
        n<- n-1
        Nd<- Nd + 1
        D[Nd] <-  t
        if (n >0)
        {
          Y <- rgamma(1,1,12)
          Td <-t + Y
        }
      }
    }
    
    #Case 4: close and no more customer
    if (min(Ta,Td) > T & n==0)
    {
      Tp<-max(t-T,0)
      Timepast3[i] <-Tp
    }
    if(min(Ta, Td) >= T) break
  }
  
  cust.wait3[i] <- sum(D-A)/Na
  depart3[i] <-t
  i <- i+1
}
mean(cust.wait3)  #average wait time for teller 3

## Taking average of the three average wait times
mean(c(cust.wait1, cust.wait2, cust.wait3))
```